Component response rate variation underlies the stability of complex systems

### 2020-03-20 notebook/ms.Rmd
Fix the notation in several places, particularly around the diagonal matrix gamma versus the vector diag(gamma)

<!--- Tempting to remove this section entirely, but if there is enough space for it, then maybe retain it; reviewers did not mention a problem with it, and it's a short and relevant point to make.  In any case, it should be swapped with the section on feasibility below. --->
**Targeted manipulation of $\mathbf{\gamma}$**. To further investigate the potential of $\sigma^{2}_{\gamma}$ to be stabilising, I used a genetic algorithm. Genetic algorithms are heuristic tools that mimic evolution by natural selection, and are useful when the space of potential solutions (in this case, possible combinations of $\gamma$ values leading to stability in a complex system) is too large to search exhaustively [@Hamblin2013]. Generations of selection on $\gamma$ value combinations to minimise $\Re(\lambda_{max})$ demonstrated the potential for $\sigma^{2}_{\gamma}$ to increase system stability. Across $S = \{2, 3, ..., 39, 40\}$, sets of $\gamma$ values were found that resulted in stable systems with probabilities that were up to four orders of magnitude higher than when $\gamma = 1$ (see Supplementary Information), meaning that stability could often be achieved by manipulating $S$ $\gamma$ values rather than $S \times S$ $\mathbf{M}$ elements (i.e., by manipulating component response rates rather than interactions between components).

@@

**Component response rate ($\mathbf{\gamma}$) variation**. In a synthesis of eco-evolutionary feedbacks on community stability, Patel et al. model a system that includes a vector of potentially changing species densities ($\mathbf{n}$) and a vector of potentially evolving traits ($\mathbf{x}$) [@Patel2018]. For any species $i$ or trait $j$, change in species density ($n_{i}$) or trait value ($x_{j}$) with time ($t$) is a function of the vectors $\mathbf{n}$ and $\mathbf{x}$,

@@

$$\frac{d\mathbf{v}(t)}{dt} = \mathbf{\gamma} \mathbf{A}\mathbf{v}(t).$$

In the above, $\mathbf{\gamma}$ is a diagonal matrix in which elements correspond to individual component response rates. Therefore, $\mathbf{M} = \mathbf{\gamma} \mathbf{A}$ defines the change in values of system components and can be analysed using the techniques of May [@May1972; @May1973; @Ahmadian2015]. In these analyses, row means of $\mathbf{A}$ are expected to be identical, but variation around this expectation will naturally arise due to random sampling of $\mathbf{A}$ off-diagonal elements and finite $S$. In simulations, the total variation in $\mathbf{M}$ row means that is attributable to $\mathbf{A}$ is small relative to that attributable to $\mathbf{\gamma}$, especially at high $S$. Variation in $\mathbf{\gamma}$ specifically isolates the effects of differing component response rates, hence causing differences in expected $\mathbf{M}$ row means.

@@

**Genetic algorithm**. Ideally, to investigate the potential of $\sigma^{2}_{\gamma}$ for increasing the proportion of stable complex systems, the search space of all possible $\gamma$ vectors would be evaluated for each unique $\mathbf{M = \gamma A}$. This is technically impossible because $\gamma_{i}$ can take any real value between 0-2, but even rounding $\gamma_{i}$ to reasonable values would result in a search space too large to practically explore. Under these conditions, genetic algorithms are highly useful tools for finding practical solutions by mimicking the process of biological evolution [@Hamblin2013]. In this case, the practical solution is finding vectors of $\mathbf{\gamma}$ that decrease the most positive real eigenvalue of $\mathbf{M}$. The genetic algorithm used achieves this by initialising a large population of 1000 different potential $\mathbf{\gamma}$ vectors and allowing this population to evolve through a process of mutation, crossover (swaping $\gamma_{i}$ values between vectors), selection, and reproduction until either a $\mathbf{\gamma}$ vector is found where all $\Re(\lambda) < 0$ or some "giving up" critiera is met.

For each $S = \{2, 3, ..., 39, 40\}$, the genetic algorithm was run for 100000 random $\mathbf{M = \gamma A}$ ($\sigma_{A} = 0.4$, $C = 1$). The genetic algorithm was initialised with a population of 1000 different $\mathbf{\gamma}$ vectors with elements sampled i.i.d from $\gamma_{i} \sim \mathcal{U}(0, 2)$. Eigenanalysis was performed on the $\mathbf{M}$ resulting from each $\mathbf{\gamma}$ vector, and the 20 $\mathbf{\gamma}$ vectors resulting in $\mathbf{M}$ with the lowest $\Re(\lambda_{max})$ each produced 50 clonal offspring with subsequent random mutation and crossover between the resulting new generation of 1000 $\mathbf{\gamma}$ vectors. Mutation of each $\gamma_{i}$ in a $\mathbf{\gamma}$ vector occurred with a probability of 0.2, resulting in a mutation effect of size $\mathcal{N}(0, 0.02)$ being added to generate the newly mutated $\gamma_{i}$ (any $\gamma_{i}$ values that mutated below zero were multiplied by $-1$, and any values that mutated above 2 were set to 2). Crossover occurred between two sets of 100 $\mathbf{\gamma}$ vectors paired in each generation; vectors were randomly sampled with replacement among but not within sets. Vector pairs selected for crossover swapped all elements between and including two $\gamma_{i}$ randomly selected with replacement (this allowed for reversal of vector element positions during crossover; e.g., $\{\gamma_{4}, \gamma_{5}, \gamma_{6}, \gamma_{7}\} \to \{\gamma_{7}, \gamma_{6}, \gamma_{5}, \gamma_{4}\}$ ). The genetic algorithm terminated if a stable $\mathbf{M}$ was found, 20 generations occurred, or if the mean $\mathbf{\gamma}$ fitness increase between generations was less than 0.01 (where fitness was defined as $W_{\gamma} = -\Re(\lambda_{max})$ for $\mathbf{M}$).

>>>>

<!--- Tempting to remove this section entirely, but if there is enough space for it, then maybe retain it; reviewers did not mention a problem with it, and it's a short and relevant point to make.  In any case, it should be swapped with the section on feasibility below. --->
**Targeted manipulation of $\boldsymbol{\gamma}$**. To further investigate the potential of $\sigma^{2}_{\gamma}$ to be stabilising, I used a genetic algorithm. Genetic algorithms are heuristic tools that mimic evolution by natural selection, and are useful when the space of potential solutions (in this case, possible combinations of $\gamma$ values leading to stability in a complex system) is too large to search exhaustively [@Hamblin2013]. Generations of selection on $\gamma$ value combinations to minimise $\Re(\lambda_{max})$ demonstrated the potential for $\sigma^{2}_{\gamma}$ to increase system stability. Across $S = \{2, 3, ..., 39, 40\}$, sets of $\gamma$ values were found that resulted in stable systems with probabilities that were up to four orders of magnitude higher than when $\gamma = 1$ (see Supplementary Information), meaning that stability could often be achieved by manipulating $S$ $\gamma$ values rather than $S \times S$ $\mathbf{M}$ elements (i.e., by manipulating component response rates rather than interactions between components).

@@

**Component response rate ($\boldsymbol{\gamma}$) variation**. In a synthesis of eco-evolutionary feedbacks on community stability, Patel et al. model a system that includes a vector of potentially changing species densities ($\mathbf{n}$) and a vector of potentially evolving traits ($\mathbf{x}$) [@Patel2018]. For any species $i$ or trait $j$, change in species density ($n_{i}$) or trait value ($x_{j}$) with time ($t$) is a function of the vectors $\mathbf{n}$ and $\mathbf{x}$,

@@

$$\frac{d\mathbf{v}(t)}{dt} = \boldsymbol{\gamma} \mathbf{A}\mathbf{v}(t).$$

In the above, $\boldsymbol{\gamma}$ is a diagonal matrix in which elements correspond to individual component response rates. Therefore, $\mathbf{M} = \boldsymbol{\gamma} \mathbf{A}$ defines the change in values of system components and can be analysed using the techniques of May [@May1972; @May1973; @Ahmadian2015]. In these analyses, row means of $\mathbf{A}$ are expected to be identical, but variation around this expectation will naturally arise due to random sampling of $\mathbf{A}$ off-diagonal elements and finite $S$. In simulations, the total variation in $\mathbf{M}$ row means that is attributable to $\mathbf{A}$ is small relative to that attributable to $\boldsymbol{\gamma}$, especially at high $S$. Variation in $\boldsymbol{\gamma}$ specifically isolates the effects of differing component response rates, hence causing differences in expected $\mathbf{M}$ row means.

@@

**Genetic algorithm**. Ideally, to investigate the potential of $\sigma^{2}_{\gamma}$ for increasing the proportion of stable complex systems, the search space of all possible $diag(\boldsymbol{\gamma})$ vectors would be evaluated for each unique $\mathbf{M = \gamma A}$. This is technically impossible because $\gamma_{i}$ can take any real value between 0-2, but even rounding $\gamma_{i}$ to reasonable values would result in a search space too large to practically explore. Under these conditions, genetic algorithms are highly useful tools for finding practical solutions by mimicking the process of biological evolution [@Hamblin2013]. In this case, the practical solution is finding vectors of $diag(\boldsymbol{\gamma})$ that decrease the most positive real eigenvalue of $\mathbf{M}$. The genetic algorithm used achieves this by initialising a large population of 1000 different potential $diag(\boldsymbol{\gamma})$ vectors and allowing this population to evolve through a process of mutation, crossover (swaping $\gamma_{i}$ values between vectors), selection, and reproduction until either a $diag(\boldsymbol{\gamma})$ vector is found where all $\Re(\lambda) < 0$ or some "giving up" critiera is met.

For each $S = \{2, 3, ..., 39, 40\}$, the genetic algorithm was run for 100000 random $\mathbf{M = \gamma A}$ ($\sigma_{A} = 0.4$, $C = 1$). The genetic algorithm was initialised with a population of 1000 different $diag(\boldsymbol{\gamma})$ vectors with elements sampled i.i.d from $\gamma_{i} \sim \mathcal{U}(0, 2)$. Eigenanalysis was performed on the $\mathbf{M}$ resulting from each $\boldsymbol{\gamma}$, and the 20 $diag(\boldsymbol{\gamma})$ vectors resulting in $\mathbf{M}$ with the lowest $\Re(\lambda_{max})$ each produced 50 clonal offspring with subsequent random mutation and crossover between the resulting new generation of 1000 $diag(\boldsymbol{\gamma})$ vectors. Mutation of each $\gamma_{i}$ in a $diag(\boldsymbol{\gamma})$ vector occurred with a probability of 0.2, resulting in a mutation effect of size $\mathcal{N}(0, 0.02)$ being added to generate the newly mutated $\gamma_{i}$ (any $\gamma_{i}$ values that mutated below zero were multiplied by $-1$, and any values that mutated above 2 were set to 2). Crossover occurred between two sets of 100 $diag(\boldsymbol{\gamma})$ vectors paired in each generation; vectors were randomly sampled with replacement among but not within sets. Vector pairs selected for crossover swapped all elements between and including two $\gamma_{i}$ randomly selected with replacement (this allowed for reversal of vector element positions during crossover; e.g., $\{\gamma_{4}, \gamma_{5}, \gamma_{6}, \gamma_{7}\} \to \{\gamma_{7}, \gamma_{6}, \gamma_{5}, \gamma_{4}\}$ ). The genetic algorithm terminated if a stable $\mathbf{M}$ was found, 20 generations occurred, or if the mean $\boldsymbol{\gamma}$ fitness increase between generations was less than 0.01 (where fitness was defined as $W_{\gamma} = -\Re(\lambda_{max})$ for $\mathbf{M}$).
